FoRTE baseline

### 2019-10-14 analysis/paper/paper.Rmd
Compile paper separately from drake
Makes it easier to work on it on different machines.

```{r setup, echo = FALSE}
library(knitr)
library(here)
library(drake)

@@

```{r param-table}
tbl <- readd(param_table)
tbl <- subset(tbl, Ran == "yes", select = c(-unit_parsed, -Ran))
tbl[["ED Name"]] <- sprintf("`%s`", tbl[["ED Name"]])
knitr::kable(tbl, escape = FALSE, caption = paste0(
  "Table 2: Parameter descriptions. ",
  "For additional information, see the ",
  "[ED2 model wiki](https://github.com/EDmodel/ED2/wiki/PFT-parameters)."
))
```

@@

For each plant functional type, we generated a distribution of parameter values via the Predictive Ecosystem Analyzer (PEcAn) trait-meta analysis [@lebauer_2013_facilitating; see also @dietze_2014_quantitative; @raczka_2018_what].
Prior distributions for this meta-analysis were largely adapted from previous ED2 parameter uncertainty studies [@dietze_2014_quantitative; @raczka_2018_what; @viskari_2019_influence], and are shown in detail in Supplementary Information X. <!-- TODO Priors supplement; maybe describe entire meta-analysis? -->
Species trait data for this meta-analysis came from existing records in the BETY database [www.betydb.org, @lebauer_2017_betydb], as well as from publicly available records in the TRY database [www.try-db.org, @kattge_2011_try] and, specifically for leaf optical properties, from Shiklomanov (2019 {dissertation, chapter 3}).
The resulting parameter distributions are shown in Figure 1.
<!-- TODO more on this? -->
All parameters not described in this section were set to their ED-2.2 PFT-specific default values.

@@

## Impacts of structure

```{r , fig.cap = structure_default_cap}
structure_default_cap <- paste(
  "Annual NPP and LAI from ED2 runs",
  "with default parameters",
  "across different configurations."
)
readd(structure_compare_default_knit)
```

@@

```{r , fig.cap = structure_default_light_cap}
structure_default_light_cap <- paste(
  "Radiation profiles of cohorts from ED2 runs with",
  "default parameters, across different configurations."
  "Each point represents a single cohort of the given PFT,"
  "position on the y-axis represents the height of that cohort,"
  "and position on the x-axis represents the relative light level",
  "received by that cohort."
)
readd(structure_default_light_knit)
```

@@

```{r param_dist, fig.cap = param_dist_cap}
param_dist_cap <- paste(
  htmlfig(),
  "Input parameter distributions from PEcAn trait meta-analysis."
)
readd(param_dist_knit)
```

`r mdcap(param_dist_cap)`

@@

```{r , fig.cap = default_median_cap}
default_median_cap <- paste(
  "Comparing results of ED2 with default parameters and",
  "posterior median parameters from trait meta-analysis",
  "across different strucures."
)
readd(default_median_fluxes_knit)
```

```{r}
default_median_cap <- paste(
  "Comparing results of ED2 PFT-level LAI",
  "with default parameters",
  "posterior median parameters from trait meta-analysis",
  "across different strucures."
)
readd(default_median_lai_knit)
```

@@

## Summary

```{r summary-ts-plot, fig.cap = summary_ts_cap}
summary_ts_cap <- paste(
  htmlfig(),
  "ED2 plot-level predictions of",
  "net (NPP) primary productivity and",
  "total leaf area index (LAI),",
  "by model configuration.",
  "The dashed lines indicate the 10th and 90th quantiles of results.",
  "Black dot with error bars is the observed mean and min/max value",
   "(TODO REF)."
)
readd(summary_ts_plot_knit)
```

`r mdcap(summary_ts_cap)`

<!-- - Structure x parameter uncertainty -->
<!--   - Figure: Time series of NPP, LAI, and relative PFT dominance -->
<!--   - Parameter uncertainty is much larger than structural uncertainty -->
<!--   - Figure: Pairs of NPP, LAI, and PFT fraction, for each structure (first draw all, then pick relevant subsets) -->
<!--   - Q: How many runs end up with the right composition (EH-dominated) at peak (~1925), ~1950, and 1999? -->

@@

Model projections of net primary productivity (NPP) and leaf area index (LAI) varied with both parameter value and model structure (Figure X).
All model configurations exhibited a similar trajectory of rapid growth in the first 15-20 years followed by a gradual decline to a lower, stable value around 1975 (with individual ensemble members sometimes undergoing sudden collapse in specific years).
The ensemble spread within any single configuration---i.e. the parameter uncertainty---was similar across all configurations; both NPP and LAI varied 5-10 fold at any given timestep depending on parameter values.
Within all configurations, most (>90%) ensemble members under-predicted observed present-day NPP and LAI by year 2000, though individual ensemble members from some configurations over-predicted one or both variables.
Predictions of peak NPP (between 1910 and 1950 for most simulations) were closer to the observed present-day NPP (though also with large ensemble spread), but predictions of LAI during this period were still too low.

```{r pft-jitterplot, fig.cap = pft_jitterplot_cap}
pft_jitterplot_cap <- paste(
  htmlfig(),
  "Fraction of total aboveground biomass (AGB)",
  "of each plant functional type (PFT)",
  "predicted by different parameter-configuration combionations",
  "of ED2."
)
readd(pft_jitterplot_knit)

```

```{r pft-ternplot, fig.cap = pft_tern_cap}
pft_tern_cap <- paste(
  htmlfig(),
  "Ternary plot of predicted PFT composition in 1920, 1950, and 1980.",
  "Ternary coordinates are as follows:",
  "L is Early Hardwood; T is the sum of Mid and Late Hardwood; R is Pine.",
  "Triangle fill color indicates the number of points within that bin."
)
readd(pft_tern_knit)
```

@@

```{r pairs-time, fig.cap = pairs_time_cap}
pairs_time_cap <- paste(
  htmlfig(),
  "ED2 predictions of annual mean NPP and growing-season LAI in 1920, 1950, and 1999",
  "by model type.",
  "Each point indicates a single ensemble run with a different set of parameters.",
  "Solid line is the best-fit generalized additive model (GAM).",
  "Dashed lines indicate the error bounds of present-day observed NPP and LAI.",
  "Text labels match the rows in Figure X below."
)
readd(npp_lai_pairs_knit)
```

@@

```{r lai-pft-plot, fig.cap = lai_pft_cap}
lai_pft_cap <- paste(
  htmlfig(),
  "PFT-level predictions of leaf area index (LAI) by model configuration and parameterization.",
  "Each row of plots is a simulation with the same set of parameters across model configurations.",
  "Letters correspond to the labels in Figure 3.",
  "Note that these are predictions of total leaf area index summed across all cohorts of the same PFT,",
  "meaning that higher LAI in one PFT does not necessarily indicate dominance, size, etc."
)
readd(lai_pft_plot_knit)
```

@@

```{r within-vs-across, fig.cap = within_vs_across_cap, fig.dim = c(9, 4)}
within_vs_across_cap <- paste(
  htmlfig(),
  "Comparison of parametric uncertainty within ensembles (colored bars, colored by model type)",
  "and \"structural\" uncertainty (variance in ensemble means; black bar) by output variable.",
  "Output variables are expressed as growing season averages for all years after 1910."
)
readd(within_across_plot_knit)
```

`r mdcap(within_vs_across_cap)`

@@

```{r uncertainty-analysis, fig.cap = uncertainty_analysis_cap}
uncertainty_analysis_cap <- paste(
  htmlfig(),
  "PEcAn-like parameter sensitivity and uncertainty analysis, by model type,",
  "for 1975-2000 averaged growing season net primary productivity (left) and leaf area index (right).",
  "Elasticity (top) is the normalized sensitivity of the model to a fixed change in the parameter.",
  "Partial variance (bottom) describes the overall contribution of the parameter to model predictive uncertainty based on the combination of parameter uncertainty and model sensitivity.",
  "Each panel shows the top 10 parameter values, in terms of the absolute value of the corresponding quantity.",
  "Input parameter distributions are shown in Figure 1.",
  "Note that x-axis scales vary across panels."
)
readd(sensitivity_plot_knit)
```

`r mdcap(uncertainty_analysis_cap)`

>>>>

```{r setup, include = FALSE}
library(knitr)
library(here)

fig <- function(...) {
  knitr::include_graphics(here::here("analysis", "figures", ...))
}

@@

```{r param-table}
tbl <- readr::read_csv(here(
  "analysis", "data", "derived-data", "parameter-table.csv"
))
tbl <- subset(tbl, Ran == "yes", select = c(-unit_parsed, -Ran))
tbl[["ED Name"]] <- sprintf("`%s`", tbl[["ED Name"]])
kable(tbl, escape = FALSE, caption = paste0(
  "Table 2: Parameter descriptions. ",
  "For additional information, see the ",
  "[ED2 model wiki](https://github.com/EDmodel/ED2/wiki/PFT-parameters)."
))
```

@@

For each plant functional type, we generated a distribution of parameter values via the Predictive Ecosystem Analyzer (PEcAn) trait-meta analysis [@lebauer_2013_facilitating; see also @dietze_2014_quantitative; @raczka_2018_what].
Prior distributions for this meta-analysis were largely adapted from previous ED2 parameter uncertainty studies [@dietze_2014_quantitative; @raczka_2018_what; @viskari_2019_influence], and are shown in detail in Supplementary Information X. <!-- TODO Priors supplement; maybe describe entire meta-analysis? -->
Species trait data for this meta-analysis came from existing records in the BETY database [www.betydb.org, @lebauer_2017_betydb], as well as from publicly available records in the TRY database [www.try-db.org, @kattge_2011_try] and, specifically for leaf optical properties, from Shiklomanov (2019 {dissertation, chapter 3}).
All parameters not described in this section were set to their ED-2.2 PFT-specific default values.

@@

## Impacts of structure alone

```{r , fig.cap = structure_default_cap}
structure_default_cap <- paste(
  htmlfig(),
  "Annual NPP and LAI from ED2 runs",
  "with default parameters",
  "across different configurations."
)
fig("structure-compare-default.png")
```

@@

```{r , fig.cap = structure_default_light_cap}
structure_default_light_cap <- paste(
  htmlfig(),
  "Radiation profiles of cohorts from ED2 runs with",
  "default parameters, across different configurations.",
  "Each point represents a single cohort of the given PFT,",
  "position on the y-axis represents the height of that cohort,",
  "and position on the x-axis represents the relative light level",
  "received by that cohort."
)
fig("default-light-levels.png")
```

@@

```{r param_dist, fig.cap = param_dist_cap}
param_dist_cap <- paste(
  htmlfig(),
  "Input parameter distributions from PEcAn trait meta-analysis."
)
fig("param-dist.png")
```

@@

```{r default-median-flux, fig.cap = default_median_flux_cap}
default_median_flux_cap <- paste(
  htmlfig(),
  "Comparing results of ED2 with default parameters and",
  "posterior median parameters from trait meta-analysis",
  "across different strucures."
)
fig("default-median-fluxes.png")
```

```{r default-median-lai , fig.cap = default_median_lai_cap}
default_median_lai_cap <- paste(
  htmlfig(),
  "Comparing results of ED2 PFT-level LAI",
  "with default parameters",
  "posterior median parameters from trait meta-analysis",
  "across different strucures."
)
fig("default-median-lai.png")
```

@@

## Structure vs. parameter uncertainty

```{r summary-ts-plot, fig.cap = summary_ts_cap}
summary_ts_cap <- paste(
  htmlfig(),
  "ED2 plot-level predictions of",
  "net (NPP) primary productivity and",
  "total leaf area index (LAI),",
  "by model configuration.",
  "The dashed lines indicate the 10th and 90th quantiles of results.",
  "Black dot with error bars is the observed mean and min/max value",
   "(TODO REF)."
)
fig("summary-ts-plot.png")
```

<!-- - Structure x parameter uncertainty -->
<!--   - Parameter uncertainty is much larger than structural uncertainty -->
<!--   - Figure: Pairs of NPP, LAI, and PFT fraction, for each structure (first draw all, then pick relevant subsets) -->
<!--   - Q: How many runs end up with the right composition (EH-dominated) at peak (~1925), ~1950, and 1999? -->

@@

Model projections of net primary productivity (NPP) and leaf area index (LAI) varied with both parameter value and model structure (Figure X).
All model configurations exhibited a similar trajectory of rapid growth in the first 15-20 years followed by a gradual decline to a lower, stable value around 1975 (with individual ensemble members sometimes undergoing sudden collapse in specific years).
The ensemble spread within any single configuration---i.e. the parameter uncertainty---was similar and large across all configurations; both NPP and LAI varied 5-10 fold at any given timestep depending on parameter values.
Within all configurations, most (>90%) ensemble members under-predicted observed present-day NPP and LAI by year 2000, though individual ensemble members from some configurations over-predicted one or both variables.
Predictions of peak NPP (between 1910 and 1950 for most simulations) were closer to the observed present-day NPP (though also with large ensemble spread), but predictions of LAI during this period were still too low.

```{r pft-jitterplot, fig.cap = pft_jitterplot_cap}
pft_jitterplot_cap <- paste(
  htmlfig(),
  "Fraction of total aboveground biomass (AGB)",
  "of each plant functional type (PFT)",
  "predicted by different parameter-configuration combionations",
  "of ED2."
)
fig("pft-jitterplot.png")
```

```{r pft-ternplot, fig.cap = pft_tern_cap}
pft_tern_cap <- paste(
  htmlfig(),
  "Ternary plot of predicted PFT composition in 1920, 1950, and 1980.",
  "Ternary coordinates are as follows:",
  "L is Early Hardwood; T is the sum of Mid and Late Hardwood; R is Pine.",
  "Triangle fill color indicates the number of points within that bin."
)
fig("pft-tern.png")
```

@@

```{r pairs-time, fig.cap = pairs_time_cap}
pairs_time_cap <- paste(
  htmlfig(),
  "ED2 predictions of annual mean NPP and growing-season LAI in 1920, 1950, and 1999",
  "by model type.",
  "Each point indicates a single ensemble run with a different set of parameters.",
  "Solid line is the best-fit generalized additive model (GAM).",
  "Dashed lines indicate the error bounds of present-day observed NPP and LAI.",
  "Text labels match the rows in Figure X below."
)
fig("npp-lai-pairs.png")
```

@@

```{r lai-pft-plot, fig.cap = lai_pft_cap}
lai_pft_cap <- paste(
  htmlfig(),
  "PFT-level predictions of leaf area index (LAI) by model configuration and parameterization.",
  "Each row of plots is a simulation with the same set of parameters across model configurations.",
  "Letters correspond to the labels in Figure 3.",
  "Note that these are predictions of total leaf area index summed across all cohorts of the same PFT,",
  "meaning that higher LAI in one PFT does not necessarily indicate dominance, size, etc."
)
fig("lai-pft.png")
```

@@

```{r within-vs-across, fig.cap = within_vs_across_cap, fig.dim = c(9, 4), eval = FALSE}
within_vs_across_cap <- paste(
  htmlfig(),
  "Comparison of parametric uncertainty within ensembles (colored bars, colored by model type)",
  "and \"structural\" uncertainty (variance in ensemble means; black bar) by output variable.",
  "Output variables are expressed as growing season averages for all years after 1910."
)
readd(within_across_plot_knit)
```

@@

```{r uncertainty-analysis, fig.cap = uncertainty_analysis_cap, eval = FALSE}
uncertainty_analysis_cap <- paste(
  htmlfig(),
  "Parameter sensitivity and uncertainty analysis, by model type,",
  "for 1975-2000 averaged growing season net primary productivity (left) and leaf area index (right).",
  "Elasticity (top) is the normalized sensitivity of the model to a fixed change in the parameter.",
  "Partial variance (bottom) describes the overall contribution of the parameter to model predictive uncertainty based on the combination of parameter uncertainty and model sensitivity.",
  "Each panel shows the top 10 parameter values, in terms of the absolute value of the corresponding quantity.",
  "Input parameter distributions are shown in Figure 1.",
  "Note that x-axis scales vary across panels."
)
readd(sensitivity_plot_knit)
```
